---
title: "Testing Residualized Centile Fan Plots"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
```

```{r}
pacman::p_load(ggplot2, tidyverse, gamlss, gamlss.ggplots, dplyr)
# install.packages("devtools")
# devtools::install_github("BGDlab/gamlssTools", force=TRUE)
# library(gamlssTools)
```

Using the iris dataset to simulate models with nonlinear effects of age and eTIV and random effect of site
```{r}
set.seed(123)  # For reproducibility

# Set sample size
n <- 100

# Generate age (range: 20-80 years)
age <- runif(n, 20, 80)

# Generate sex (0 = female, 1 = male)
# Making this independent for now
sex <- rbinom(n, 1, 0.5)

site <- sample(c("A", "B", "C"), n, replace=TRUE)

# Generate eTIV (estimated Total Intracranial Volume)
# Positively correlated with age
# Range approximately 1200-1800 cm³
# Males tend to have slightly larger eTIV
eTIV <- 1400 + 0.3 * age + 100 * sex + rnorm(n, 0, 50)

# Generate phenotype score (positively correlated with both age and eTIV)
# This could represent some clinical or cognitive measure
pheno <- 50 + 0.4 * age + 0.02 * eTIV + rnorm(n, 0, 5)

# Combine into a data frame
df <- data.frame(
  id = 1:n,
  sex = sex,
  age = age,
  eTIV = eTIV,
  site = as.factor(site),
  pheno = pheno
)
```

```{r}
test_model <- gamlss(formula = pheno ~ pb(age) + random(site) + pb(eTIV), 
                     sigma.formula = ~ age + eTIV, 
                     family = BCTo, data = df)
```

#Test with current gamlssTools functions

```{r}
make_centile_fan(test_model, df, x_var="age", color_var="site") 
make_centile_fan(test_model, df, x_var="age", remove_point_effect = c("site", "eTIV"))
make_centile_fan(test_model, df, x_var="age", zero_effect = c("eTIV"))
#sim_data(df, "age", "site", test_model, zero_effect = "eTIV")

sim_df <- sim_data(df, "age", "site", test_model, zero_effect="eTIV")
make_centile_fan(test_model, df, x_var="age", sim_data_list = sim_df)
```

try residualizing eTIV from points
```{r}
ggplot(df) +
  geom_point(aes(x=age, y=pheno, color=site))
make_centile_fan(test_model3, df, x_var="age", show_points = TRUE, 
                 #remove_point_effect = c("eTIV", "site"),
                 #remove_cent_effect = c("site", "eTIV")
                 )

make_centile_fan(test_model3, df, x_var="age", show_points = TRUE, 
                 remove_point_effect = c("eTIV"),
                 #remove_cent_effect = c("site", "eTIV")
                 )

make_centile_fan(test_model3, df, x_var="age", show_points = TRUE, 
                 remove_point_effect = c("eTIV", "site"),
                 #remove_cent_effect = c("site")
                 )
cent_cdf(test_model, df, interval_var="age", n=4, plot = FALSE)

make_centile_fan(test_model, df, x_var="eTIV", show_points = TRUE, 
                 remove_point_effect = c("age", "site"),
                 #remove_cent_effect = "site"
                 )
cent_cdf(test_model, df, interval_var="eTIV", n=4, plot=FALSE)

#resid_data(test_model, df, rm_terms="eTIV")
```


still off - try residualizing out site effects (removing from centile fans as well, because centile lines are simulated at level=mode while Betas are calculated against whatever gamlss sets as the base level, which may or may not be the same)

```{r}
make_centile_fan(test_model, df, x_var="age", 
                 remove_cent_effect = c("site"),
                 remove_point_effect = c("site", "eTIV"))
```

#Other Ideas
what if i just predict df from the gamlss model, except i replace the values of df i'm residualizing with the means?

```{r}
test_model4 <- gamlss(formula = pheno ~ pb(age) + site, 
                     sigma.formula = ~ age, 
                     family = GG, data = df)

make_centile_fan(test_model4, x_var = "age", df)

make_centile_fan(test_model4, x_var = "age", df, remove_cent_effect = "site")

centile_fan_resid(test_model4, df, x_var="age", resid_effect = "site")
```

```{r}
df2 <- df %>%
  mutate(site = mode(df$site))

test_pred <- predictAll(test_model4, newdata = df2, data=df)

test_pred2 <- predictAll(test_model4, newdata = df, data=df)

test_pheno <- df$pheno + (test_pred$mu - test_pred2$mu)

ggplot() +
  geom_point(aes(x=df$age, y=test_pheno))

resid_data(test_model4, df, og_data=df, rm_terms="site")
```


Currently, the subfunction that removes gamlss effects from points (`resid_data()`) only removes mu effects. Will try refitting the model without eTIV in sigma to see if this aligns better.

```{r}
test_model2 <- gamlss(formula = pheno ~ pb(age) + random(site) + 
    pb(eTIV), sigma.formula = ~ age, 
    family = GG, data = df)

make_centile_fan(test_model2, df, x_var="age", 
                 remove_cent_effect = "site",
                 remove_point_effect = c("eTIV", "site"))
```
doesn't seem to help

what if eTIV is linear?
```{r}
test_model3 <- gamlss(formula = pheno ~ pb(age) + random(site) + eTIV, sigma.formula = ~ age, 
    family = GG, data = df)

make_centile_fan(test_model3, df, x_var="age", 
                 remove_cent_effect = "site",
                 remove_point_effect = "site")
make_centile_fan(test_model3, df, x_var="age", 
                 remove_cent_effect = "site",
                 remove_point_effect = c("eTIV", "site"))
```

looks much better!

```{r}
test_model4 <- gamlss(formula = pheno ~ pb(age) + site + eTIV, 
    family = NO, data = df)

make_centile_fan(test_model4, df, x_var="age",
                 color_var="site", average_over=TRUE,
                 remove_point_effect = c("site", "eTIV"))
```
Seems to work better with linear effects/when family=NO, trying resisdualizing from other moments

```{r}
make_centile_fan(test_model, df, x_var="age", remove_point_effect = c("site", "eTIV"))

resid_pts <- resid_data(test_model, df, df, rm_terms="eTIV")
ggplot(resid_pts) +
  geom_point(aes(x=age, y=pheno))
```

```{r}
make_centile_fan(test_model2, df, x_var="age", remove_point_effect = c("site", "eTIV"))

resid_pts <- resid_data(test_model2, df, df, rm_terms="eTIV")
ggplot(resid_pts) +
  geom_point(aes(x=age, y=pheno))
```

seems to be less wonky with NO()/removing effect in sigma

back to thinking the problem is with pb effects :(

```{r}
test_model5 <- gamlss(formula = pheno ~ pb(age) + site + pb(eTIV), 
    family = NO, data = df)


effects_link <- predict(object = test_model5,
                             newdata = df,
                             what = "mu",
                             data = df,
                             type="terms"
                            )
new_df <- data.frame(
  age = mean(df$age),
  eTIV=mean(df$eTIV),
  site="setosa"
)

mean_link <- predict(object = test_model5,
                             newdata = new_df,
                             what = "mu",
                             data = df,
                             type="terms"
                            )


df_test <- df
eff_link <- effects_link %>%
  as.data.frame()
eff_link[['pb(eTIV)']]

df_test$link_effects <- test_model5$mu.coefSmo[[2]]$fv
  #eff_link[['pb(eTIV)']]
  
df_test <- df_test %>%
  mutate(pheno_minus = pheno - link_effects,
         pheno_plus = pheno + link_effects,
         linear_effect = eTIV * 0.5451) %>%
  mutate(pheno_minus_linear = pheno_minus + linear_effect)

# df_test$response <- predict(object = test_model5,
#                              newdata = df,
#                              what = "mu",
#                              data = df,
#                              type="response"
#                             )
# 
# predictAll(object = test_model5, newdata = df, data = df)

ggplot(df_test) +
  geom_point(aes(x=age, y=pheno_minus))
ggplot(df_test) +
  geom_point(aes(x=age, y=pheno_plus))

ggplot(df_test) +
  geom_point(aes(x=age, y=pheno_minus_linear))

make_centile_fan(test_model5, df, "age")
make_centile_fan(test_model5, df, "age", remove_point_effect=c("eTIV"))
make_centile_fan(test_model5, df, "age", "site", average_over=TRUE, remove_point_effect=c("eTIV", "site"))
#make_centile_fan(test_model5, df, "age", remove_point_effect=c("eTIV", "site"), remove_cent_effect = "site")
```

adding effect of pb(mean(eTIV)) doesn't work :(

```{r}
getSmo(test_model5, which=2)$
```