---
title: "Model Diagnostics & Centile Scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Diagnostics & Centile Scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  markup = FALSE,
  fig.width = 8,
  fig.height = 7
)

has_charts <- requireNamespace("gamlss2charts", quietly = TRUE)
```

```{r setup}
library(gamlssTools)
```

This vignette covers checking how well a model fits and getting centile/z-score values out of it.

We'll use this toy dataset:

```{r}
set.seed(1234)
n <- 1000

# Age sampled uniformly on the log scale (birth = 280 days post-conception)
logAge <- runif(n, log10(196), log10(280 + 100 * 365.25))  # 28wk gestation to 100y
df <- data.frame(
  Age    = round(10^logAge - 280),   # days post-birth
  logAge = logAge,
  Sex    = factor(sample(c("Female", "Male"), n, replace = TRUE))
)

# Study assignment correlated with age
study_center <- c(Study_A = 2.55, Study_B = 3.15, Study_C = 3.70,
                  Study_D = 4.15, Study_E = 4.45)
study_width  <- 0.5                       # larger = more overlap between studies
study_size   <- c(1, 1, 0.8, 1.3, 0.6)     # relative recruitment volume

w <- outer(df$logAge, study_center, dnorm, sd = study_width)
w <- sweep(w, 2, study_size, "*")
df$Study <- factor(colnames(w)[apply(w, 1, function(p) sample.int(length(p), 1, prob = p))],
                   levels = names(study_center))

# Median trajectory (arbitrary volume-like units): fast perinatal rise,
# childhood peak, gradual decline through adulthood
growth  <- plogis((df$logAge - 2.70) / 0.22)
atrophy <- 1 - 0.38 * plogis((df$logAge - 3.85) / 0.32)
mu <- 900 * growth * atrophy

# Sex: males ~10% larger, with a small age interaction
mu <- mu * ifelse(df$Sex == "Male", 1.10 * (1 + 0.04 * (df$logAge - 3.5)), 1)

# Study/scanner offsets
mu <- mu * c(Study_A = 1.00, Study_B = 0.92, Study_C = 1.10,
             Study_D = 0.96, Study_E = 1.06)[df$Study]

# Multiplicative noise, wider in infancy -> heteroscedastic and right-skewed
sigma <- 0.05 + 0.07 * plogis(-(df$logAge - 3.0) / 0.35)
df$Pheno <- mu * exp(rnorm(n, 0, sigma))
```

And fit a basic model
```{r}
pheno_model <- gamlss(formula = Pheno ~ pb(logAge) + Sex + random(Study), 
                      sigma.formula= ~ pb(logAge) + Sex, data = df, family=BCCG)
```

## Model Diagnostics

There are several quantitative and qualitative checks you'll likely want to perform on your models.

First, we can return BIC and AIC built-into the gamlss model object (note: will be different when using gamlss2 objects)
```{r}
pheno_model$aic #AIC
pheno_model$sbc #BIC
```

Another helpful check is the worm plot. `wp.taki()` enhances the built-in `gamlss::wp()` by allowing you to break up plots along chunks of a continuous variable (like age) or by levels of a categorical variable (like sex), so you can make sure the model fits all groups well.

```{r}
#plot by age
wp.taki(xvar=df$logAge, resid=resid(pheno_model))

#plot by sex
df$Sex <- as.factor(df$Sex)
wp.taki(xvar=df$Sex, resid=resid(pheno_model))
```

You can also check how well the predicted centiles for each point/subject line up with the percentiles themselves - that is, about 5% of subjects' centiles should be <= 0.05.

```{r}
centile_coverage(pheno_model, df)

#we can also view this by sex or age
centile_coverage(pheno_model, df, group="Sex")
centile_coverage(pheno_model, df, interval_var = "logAge", n=4)
```

Each of those calls re-scores every subject. If you've already calculated centiles (or scoring is
slow), hand them to `centile_coverage()` directly with the `centiles` argument instead of a model:

```{r}
df$centile <- score_centiles(pheno_model, df)

centile_coverage(data=df, centiles="centiles", group="Sex")
```

`centiles` also accepts a plain numeric vector, or the dataframe returned by
`score_centiles(..., standardize=TRUE)`. `data` is only needed for grouping - with no `group` or
`interval_var`, the centiles alone are enough (`centile_coverage(centiles=df$centile)`).

## Calculating Reference Scores

Now let's say you want to get the centile values for each subject in your dataset or a new dataset.

We'll start by simulating a new, small dataset.
```{r}
set.seed(5678)
n_test <- 100

study_center_new <- c(Study_A = 2.55, Study_B = 3.15, Study_New = 4)
study_size_new   <- c(1, 1, 1)   # one weight per study above

test_logAge <- runif(n_test, log10(196), log10(280 + 100 * 365.25))
test_new <- data.frame(
  Age    = round(10^test_logAge - 280),
  logAge = test_logAge,
  Sex    = factor(sample(c("Female", "Male"), n_test, replace = TRUE)),
  # a clinical variable the model knows nothing about
  dx     = factor(sample(c("control", "case"), n_test, replace = TRUE),
                  levels = c("control", "case"))
)

w_test <- outer(test_new$logAge, study_center_new, dnorm, sd = study_width)
w_test <- sweep(w_test, 2, study_size_new, "*")
test_new$Study <- factor(colnames(w_test)[apply(w_test, 1, function(p) sample.int(length(p), 1, prob = p))],
                        levels = names(study_center_new))

test_mu <- 900 * plogis((test_new$logAge - 2.70) / 0.22) *
  (1 - 0.38 * plogis((test_new$logAge - 3.85) / 0.32))
test_mu <- test_mu * ifelse(test_new$Sex == "Male", 1.10 * (1 + 0.04 * (test_new$logAge - 3.5)), 1)
test_mu <- test_mu * c(Study_A = 1.00, Study_B = 0.92, Study_New = 0.89)[test_new$Study]

# add disease effect
test_mu <- test_mu * ifelse(test_new$dx == "case", 0.93, 1)

test_sigma <- 0.05 + 0.07 * plogis(-(test_new$logAge - 3.0) / 0.35)
test_new$Pheno <- test_mu * exp(rnorm(n_test, 0, test_sigma))

#for demo simplicity, we'll save a couple subsets
test_control <- test_new %>% filter(dx=="control")
test_df <- test_control %>% filter(Study != "Study_New")
```

Now we can return the centile values for each subject, as estimated by our model above.

```{r}
test_df$centile <- score_centiles(pheno_model, test_df)
range(test_df$centile)
```

You can check the centile score's coverage (i.e., are 10% of subjects in the 10th percentile or less?) using `centile_coverage()`

```{r}
centile_coverage(data=test_df, centiles="centiles", group="Study")
```

You can also return standardized scores, or "pseudo-z-scores"

```{r}
score_centiles(pheno_model, test_df, standardize=TRUE)
```

The same code will work to calculate centiles for the original data the model was fit on.

However, this training data isn't necessary for running `score_centiles()` on new data, which is useful when the model is saved and reloaded and/or when the training data can't be shared.

### Out-of-Sample Centile Scoring

We can also get scores for data from a new site ('Site_New') that wasn't part of the training data our model was fit on.

```{r}
table(test_control$Study)
```


Out-of-sample scoring is handled by the optional dependency, gamlss2charts, which you can install
with `remotes::install_github("andy1764/gamlss2charts@dev")`. Everything above works without it;
only the `batch_term` and `ref_data` chunks below need it.

```{r, echo = FALSE, results = "asis", eval = !has_charts}
cat("> **gamlss2charts is not installed**, so the remaining chunks are shown but not run.\n",
    "> Install it with `remotes::install_github(\"andy1764/gamlss2charts@dev\")` and rebuild\n",
    "> this vignette to see their output.\n", sep = "")
```

If we just stick with the defaults, site effects for our new site will be set to 0, which doesn't produce very good estimates.

```{r}
test_control$centile <- score_centiles(pheno_model, test_control)
centile_coverage(data=test_control, centiles="centiles", group="Study")
```

Passing the `batch_term` arg enables batch-effect estimation and removal via `gamlss2charts::predict_score()`

```{r, eval = has_charts}
#get centiles 
test_control$centile2 <- score_centiles(pheno_model, test_control, batch_term="Study")
hist(test_control$centile2)

#check how they look
centile_coverage(data=test_control, centiles="centiles2", group="Study")
```

Much better!

We can even specify a subset of the data to be used as a reference, from which batch effects are calculated.
For example, while you may assume that the controls for your new site should have centiles that are evenly distributed across the range of 0-1. However, this may not be the case with individuals who have psychopatholgy, who may have deviations you want to preserve in their centile scores.

```{r}
table(test_new$Study, test_new$dx)
```

You can use your controls to estimate and remove site effects from new batches by using the `ref_data` argument, 
which accepts a formula/character string defining how reference data should be subsetted from your `data`...

```{r, eval = has_charts}
test_new$centile <- score_centiles(pheno_model, test_new, batch_term="Study", ref_data= ~ dx=="control")

ggplot(test_new) +
  geom_histogram(aes(x=centile, fill=dx), position="dodge", alpha=.6) +
  theme_minimal()
```

... Or an entirely new dataframe, which will just be used to calculate batch effects for `data` but not
scored itself (unless these subjects are also part of your `data` input)

```{r, eval = has_charts}
score_centiles(pheno_model, test_new, batch_term="Study", ref_data=test_control)
```

Current testing suggests you need about *75 reference subjects per batch* for good (mean error ~ 1 centile) out-of-sample estimates.